Compositional Verification of Concurrent C Programs with Search Structure Templates

### 2023-04-28 concurrency/CONCUR 2023/paper.tex
intro edits

%don't say ``real'' so much
\begin{abstract}
%a bit more of a lead-in?
Concurrent search structure templates are a technique for separating the verification of a concurrent data structure into concurrency-control and data-structure components, which can then be modularly combined with no additional proof effort. In this paper, we implement the template approach in the Verified Software Toolchain, and use it to prove correctness of C implementations of fine-grained concurrent data structures. In the process, we explore some of the practical requirements and limitations of the template approach on real code. %What's our conclusion? Are the template proofs easier/fewer lines than the proofs without the templates?
We find that the template approach is generally effective on real C implementations, although some templates require modifications to the target data structure, and some data structures have operations that are not easily decomposed. This work represents a step towards practical compositional verification of real-world concurrent data structures.
\end{abstract}

\section{Introduction}
\label{sec:introduction}
Krishna et al.~proposed concurrent search structure templates~\cite{templates} as a method for proving the correctness of concurrent data structures \emph{compositionally}, separating the proof of a concurrent access method (e.g., optimistic concurrency, hand-over-hand locking, forwarding via internal links) from the proof of the underlying data structure (e.g., linked list, hashtable, B-tree). The concurrency ``templates'' are verified parametrically over data structure operations, and the data structure operations are verified without any reference to concurrency. In theory, this could allow us to prove the correctness of $n$ (single-threaded) data structure implementations and $m$ concurrency patterns, and immediately obtain $n \times m$ verified concurrent data structures. In practice, the story is more complicated: certain patterns work only for specific data structures or require the data structures to store extra information, while some internal data structure operations may not fit the template model. \wm{Maybe say more about the idea that they only built a proof of concept, and there are still important details not worked out/only tested on ad-hoc examples.}

@@

Our specific contributions are:
\begin{itemize}
\item We reimplement the template approach without flow interfaces, with a simple interface involving only the concept of ``keys belonging in this node/subtree''.
\item We implement the template approach in VST, allowing us to apply it to real C programs and obtain end-to-end correctness proofs in a single verification system.
\item To the best of our knowledge, this is the first mechanized verification of a template approach to concurrent data structure implementations in a real programming language.
\item In the process, we encounter some of the limitations of the template approach on tree data structures. \wm{It would be nice if we had a possible solution!}
\end{itemize}
%\wm{This will be useful for contributions: Than, what would you say the hardest part was in completing the proofs?}

%\than{The most challenging aspect of the project was the proofs involving traversal. It took me a considerable amount of time to understand the pre- and post-conditions, as well as determining the appropriate loop invariants. I found myself frequently switching between proving the \texttt{insertOp} and \texttt{traverse} functions in order to combine all elements. However, once I had completed the \texttt{traverse} proofs, everything became much easier.}

>>>>

%don't say ``real'' so much
\begin{abstract}
%a bit more of a lead-in?
Concurrent search structure templates are a technique for separating the verification of a concurrent data structure into concurrency-control and data-structure components, which can then be modularly combined with no additional proof effort. In this paper, we implement the template approach in the Verified Software Toolchain (VST), and use it to prove correctness of C implementations of fine-grained concurrent data structures. This involves translating code, specifications, and proofs to the idiom of C and VST, and gives us another look at the requirements and limitations of the template approach.
We encounter several questions about the boundaries between template and data structure, as well as some common data structure operations that cannot naturally be decomposed into templates. Nonetheless, the approach appears promising for modular verification of real-world concurrent data structures.
\end{abstract}

\section{Introduction}
\label{sec:introduction}
Krishna et al.~proposed concurrent search structure templates~\cite{templates} as a method for proving the correctness of concurrent data structures \emph{compositionally}, separating the proof of a concurrent access method (e.g., optimistic concurrency, hand-over-hand locking, forwarding via internal links) from the proof of the underlying data structure (e.g., linked list, hashtable, B-tree). The concurrency ``templates'' are verified parametrically over data structure operations, and the data structure operations are verified without any reference to concurrency. In theory, this could allow us to prove the correctness of $n$ (single-threaded) data structure implementations and $m$ concurrency patterns, and immediately obtain $n \times m$ verified concurrent data structures. In practice, the story is more complicated: certain patterns work only for specific data structures or require the data structures to store extra information, while some internal data structure operations may not fit the template model. %\wm{Maybe say more about the idea that they only built a proof of concept, and there are still important details not worked out/only tested on ad-hoc examples.}

@@

Our specific contributions are:
\begin{itemize}
\item We reimplement the template approach without flow interfaces, with a simple interface involving only the concept of ``keys belonging in this node/subtree''.
\item We implement the template approach in VST, allowing us to apply it to C programs and obtain end-to-end correctness proofs in a single verification system.
\item To the best of our knowledge, this is the first mechanized verification of a template approach to concurrent data structure implementations in a real programming language, and its first application to data structures not written specifically as case studies.
\item We highlight several questions about the boundaries of the template/data structure division, and describe a limitation of the template approach on tree data structures.
\end{itemize}
